---
title: "Homework 3"
output:
  html_document: default
  pdf_document: default
date: 'Deadline: 9 February, 12:00'
subtitle: 'Linguistic Data: Quantitative Analysis and Visualisation'
---

The solutions should be submitted via GitHub.  

## Part 1. A preliminary training  

*Do not use R (RStudio) to solve problems in Part 1. Answers won't be evaluated. *

### Problem 1

Look at the following histogram and answer the questions.

```{r, echo=FALSE}
set.seed(1234)
x <- rnorm(20, mean=3, sd=6)
hist(x, col="tomato", main="")
```

a. What is the proportion of values in the sample that exceed $5$? Explain your answer.
b. Indicate the interval where the median of this sample can lie. Explain your answer.
c. How the histogram will change if we add an element $7$ to the sample? Explain your answer.

Answers:  
a. There are 4 values lying between 5 and 10, 1 value between 15 and 20; so there are 5 out of 20 values exceeding 5  
b. The median should be somewhere between -5 and 0 because we have 11 elements less than 0, 9 - more; so the median is a mean of 2 values from -5 and 0 (which will still be in that interval)  
c. The fourth column would become taller by 1  

### Problem 2

a. Look at the histograms of two samples. They illustrate the distribution of normalized average reaction time to frequent words (in ms) in two groups of people.

```{r, echo=FALSE, fig.height=3}
set.seed(1234)
sample1 <- rnorm(100, 500, 50)
sample2 <- rnorm(100, 500, 20)
hist(sample1, col="lightblue", xlim=c(100, 900))
```

```{r, echo=FALSE, fig.height=3}
hist(sample2, col="lightgreen", xlim=c(100, 900))
```

Which of the samples has a larger variance? Explain your answer.

Answer:  
The first sample has a larger variance because its values lie in a larger interval - from 260 to 640 vs from 440 to 580; also mostly all of the values of sample 2 lie somewhere near the mean  
\newpage

b. Look at the histograms of two samples.

```{r, echo=FALSE, fig.height=8}
set.seed(12)
sample1 <- rnorm(100, 3, 5)
sample2 <- rnorm(100, 10, 0.8)
par(mfrow=c(2,2))
hist(sample1, col="lightblue", xlim=c(-40, 40))
hist(sample2, col="lightgreen", xlim=c(5, 15))
```

Which of the samples has a larger variance? Explain your answer.

Answer:  
The first sample has a larger variance because its values lie in a larger interval; the samples look like nearly the same though there are extra columns on the sides of the first sample which make the difference  

## Part 2

*Do not use R (RStudio) to solve problems in Part 2. Answers for problem 3 will be evaluated. Please paste YES or NO into (empty) code blocks and explain you answer below the block.*

### Problem 3

Below is the histogram of the number of mistakes students made while writing an examination essay in English. Look at the histogram and answer the questions.

```{r, echo=FALSE, fig.height=4}
set.seed(2)
mistakes <- rbeta(1000, 2, 7)*100
hist(mistakes, breaks=60, col= "deepskyblue", ylim=c(0, 40))
```

### 3.1
Is it true that $50$% students made more than $35$ mistakes?
```
NO
```
Explain your answer below:  
The columns for less than ~35 mistakes are much higher so there are a lot more than 50% of people making less than 35 mistakes  

### 3.2
Is it true that most students made no more than $10$ mistakes?
```
NO
```
Explain your answer below:  
Columns between 10 and 20 make higher sum than the columns between 0 and 10, and even more is added to that sum to represent poeple making more than 20 mistakes  

### 3.3
Which of the following values is closer to be the median of `mistakes`: $10$, $20$, $30$, $40$?
```
20
```
Explain your answer below:  
The highest columns are closer to 20 mistakes, the other columns around 20 are also quite high, so the median should be there  

### Problem 4. Exact binomial test

The null hypothesis is that $p=0$ (i.e. no success is possible). In a dataset, there is only one success out of 1 000 000 observations. Will you reject the null hypothesis? 
```
Yes, i will reject it
```
Explain your answer below:
If we make an exact binomail test - C(n,k) * 0 ^ k * 1 ^ (n-k) = 0. That makes p-value = 0, which is less than 5%, so we reject the null hypothesis

## Part 3

*Use R (RStudio) to solve problems in Part 3. Your answers will be evaluated. Please paste R code into R code blocks and explain you answer below the block, if needed. *

### Problem 5 

Here is a sample of respondents' age:  

$44$, $50$, $42$, $64$, $66$, $42$, $72$, $56$, $72$, $54$, $46$, $48$, $48$, $52$, $50$, $66$, $84$.

### 5.1
Arrange them in a vector and call it `age`. 
```{r}
age <- c(44,50,42,64,66,42,72,56,72,54,46,48,48,52,50,66,84)
```

### 5.2
Examine the type of `age` variable (numeric, character, etc).
```
double
```

### 5.3
Plot the histogram of the vector `age` with $5$ bins. Change its color to any you want. (Use either R basic or ggplot2 style for plotting.)
```{r}
hist(age, breaks = 5, col = "pink")
```


### Problem 6

Here is a series of words:  
*pie, bar, bar, pie, pie, bar, bar, chart*.

### 6.1
Arrange elements above in a vector and call it `words` 
```{r}
words <- c("pie", "bar", "bar", "pie", "pie", "bar", "bar", "chart")
```

### 6.2
Calculate the relative frequences of values in `words` measured in percent.
```{r}
words <- c("pie", "bar", "bar", "pie", "pie", "bar", "bar", "chart")
mytable <- table(words)
prop.table(mytable)
```

## Problem 7. Position of verbs in verses

The dataset [“The last words in verses”](https://raw.githubusercontent.com/LingData2019/LingData2020/master/data/poetry_last_in_lines.csv) is based on texts written in the 1820s and 1920s (Corpus of Russian Poetry of the Russian National Corpus). Authors collected only one line per author to keep observations as independent from each other as possible. 

Variables:  
- Decade — decade of creation: 1820s, 1920s.  
- RhymedNwords — the number of words in the rhyming position (usually one word, but there are two words in cases such as _вина бы_ 'I would like to get) wine' (which is rhymed with _жабы_ 'toad', see http://russian-poetry.ru/Poem.php?PoemId=18261)).  
- RhymedNsyl — the number of syllables in the rhyming position.  
- UPoS — part of speech of the last word. 
- LineText — a sampled verse.  
- Author — the author of the text.  

Can we decide that in verses written in 1920s, verbs in the rhyming posision are used differently (more often or less often) than expected for verbs in general? 

Let's assume that the probability for verbs to be used in any position ('in general') is 17% (according to [](http://www.ruscorpora.ru/new/corpora-stat.html) ).

### 7.1 State hypothesis

What is your null hypothesis $H_0$ and what is the alternative hypothesis $H_1$?
```
H0: p(verb) in 1920s was the same from general usage of verbs
H1: p(verb) in 1920s was different from the general usage of verbs
```

### 7.2

Read the dataset [“The last words in verses”](https://raw.githubusercontent.com/LingData2019/LingData2020/master/data/poetry_last_in_lines.csv). Filter out the relevant observations from 1920s, calculate the number of verbs observed in the sample, and the sample size. 
```{r}
library(dplyr)
df <- read.csv(sep = "\t", "https://raw.githubusercontent.com/LingData2019/LingData2020/master/data/poetry_last_in_lines.csv")
filtered_1920 <- df %>% 
  filter(Decade == "1920s")
verbs_1920 <- filtered_1920 %>% 
  filter(UPoS == "VERB")
observed_size_1920 <- length(verbs_1920$Decade)
sample_size <- length(df$Decade)
```

### 7.3

Use an exact binomial test to calculate p-value. 
```{r}
binom.test(observed_size_1920, sample_size, p=0.17)
```

### 7.4 Interpret results

Give your interpretation of obtained p-value. Answer the initial question: Can we decide that in verses written in 1920s, verbs are used in the rhyming posision more often or less often than expected?
```
The null hypothesis was wrong - p-value is less than 5%
```

### 7.5 
*(A bonus problem, extra points in evaluation)*.
Repeat 2.3 for verses written in the 1820s.
```{r}
filtered_1820 <- df %>% 
  filter(Decade == "1820s")
verbs_1820 <- filtered_1820 %>% 
  filter(UPoS == "VERB")
observed_size_1820 <- length(verbs_1820$Decade)

binom.test(observed_size_1820, sample_size, p=0.17)
```

Write down your general conlusions about data provided for both 1920s and 1820s data.
```
Seems like verbs were rarely used in the rhyming position long ago though then people started to use them, maybe someone started the tendency in the 1920s. It's worth mentioning that the structure of poetry and prose is quiet different and corpora has more prose - this fact may have influenced the general probability.
```

### Problem 8. One-sample t-test
Using Icelandic data on vowel duration from seminar [Link](https://raw.githubusercontent.com/LingData2019/LingData2020/master/data/icelandic.csv) test the null hypothesis that the population mean of vowel duration in speaker `shg05` equals 73 (ms).
To perform a one-sample t-test, you can use the following example of R code: 
``` 
t.test(sample, mu = 7725) # mu is a population mean
```

### 8.1
Write down a two-tailed alternative hypothesis.
```
H1: the population mean of vowel duration in speaker `shg05` not equals 73 (ms)
```

### 8.2
Perform a one-sample t-test. 
```{r}
icelandic <- read.csv("https://raw.githubusercontent.com/LingData2019/LingData2020/master/data/icelandic.csv")
filtered <- icelandic %>% 
  filter(speaker == "shg05")
t.test(filtered$vowel.dur, mu = 73)
```

### 8.3
Interpret results. 
```
Assuming p-value is more than 5% we can say that the null hypothesis was correct
```

#### Supplementary reading
Use of exact binomial test in lingiistic research:  

* Gries, Stefan Th. "Phonological similarity in multi-word units." Cognitive Linguistics 22.3 (2011): 491-510. [Link](http://citeseerx.ist.psu.edu/viewdoc/download?doi=10.1.1.186.7412&rep=rep1&type=pdf)  
Stefan Gries proves that alliteration is observed in multi-word expressions more often than in general.

* Harald Bayen (2008: 51-52) evaluates the probability of observing exactly one occurrence of the word _hare_ in the corpus sample of 1 mln words given its estimated frequency of 8.23 words per million according to the SELEX frequency database.

On measures of central tendency: 

* Levshina 2015, Chapter 3 (p. 48); Gries 2009, Chapter 1.3 (p. 116). 

On t-test: 

* Gries 2009, Chapter 3 (p. 198). 